Treebeard : An Optimizing Compiler for Tree Based ML Inference

### 2022-04-15 docs/MICRO 2022/overview.tex
overview HIR

\section{Compiler Overview}
Figure \ref{Fig:CompilerStructure} shows the high level structure of Treebeard. 
The input to TreeBeard is a JSON serialized decision forest model, it supports popular frameworks like XGBoost and LightGBM and is extensible to other frameworks.
Given an input model our compiler generates optimized inference code. Specifically it generates a callable batch inference function \texttt{predictForest} that given an array of inputs, outputs an array of model predictions.

TreeBeard specializes the code generated for inference by progressively optimizing and lowering a high level representation of the \texttt{predictForest} function down to LLVM IR \cite{LLVM}.
It is implemented as a dialect in the MLIR framework \cite{MLIR},
%Treebeard first constructs a high level representation of the decision forest inference operation. 
 At this level (HIR) it applies transformations that determine how inputs traverse the decision forest. It re-order trees in the forest and determines the high level loop structure (process one tree at a time or one input at a time). It then lowers to a traditional loop based IR (MIR). Several loop level optimizations like unrolling and pipelining are applied at this level. At the lowest level of IR TreeBeard applies layout optimizations and generates parallel code in LLVM IR. 
LLVM is then used to JIT compile the generated IR to executable code. The following sections describe each level of the IR and their lowering in more detail.

@@

\TODO{Should we describe the dialect's type system?}
\TODO{Kr : consider focusing on the example instead of verbose description of optimizations. This section can be short, descriptions can come in later sections.}
\subsection{High Level IR}
Treebeard parses the input JSON file and generates a function with a single MLIR operation, \texttt{predictForest} that represents inference using the input model on a set of rows. The operation contains within it a tree based representation of the model that can be manipulated by optimizing transformations. Transformations on the model such as tiling, tree reordering and leaf padding are done at this level. The structure of the loop nest to walk the iteration space of trees and inputs is also decided at this level of the IR. \TODO{Should we mention that there is a scheduling language to decide this?}

@@

\subsection{Mid Level IR}
The Mid Level IR makes the loop structures and tree walks more explicit. Firstly, the order in which the iteration space of trees and inputs is walked is explicitly specified in the IR through loop nests. Also, operations such as \texttt{isLeaf}, \texttt{traverseTile}, \texttt{getLeafValue} are introduced so that the traversal of trees explicitly represented. The following listing shows the IR for inference using a model with four trees on an input batch with two rows. The listed IR walks all trees for one input row before moving to the next row. One important point to note here is that details such as the data structure used for the trees are not explicitly encoded in the IR. This allows us to reuse optimization and lowering passes on this level of the IR regardless of what the final in memory representation of the model is.

>>>>

\section{Compiler Overview}
Figure \ref{Fig:CompilerStructure} shows the high level structure of Treebeard. 
The input to Treebeard is a JSON serialized decision forest model, it supports popular frameworks like XGBoost and LightGBM and is extensible to other frameworks.
Given an input model our compiler generates optimized inference code. Specifically it generates a callable batch inference function \texttt{predictForest} that given an array of inputs, outputs an array of model predictions.

Treebeard specializes the code generated for inference by progressively optimizing and lowering a high level representation of the \texttt{predictForest} function down to LLVM IR \cite{LLVM}.
%Treebeard first constructs a high level representation of the decision forest inference operation. 
At the highest level (HIR) it tiles nodes together to transform a binary tree into an $n$-ary tree and applies transformations that determine how inputs traverse the decision forest. It then lowers to a mid-level IR (MIR) where loop optimizations like loop-reordering and unrolling are applied. The compiler also figures out which loop nest to parallelize. At the lowest level of IR Treebeard applies layout optimizations and generates vectorized code to take advantage of SIMD instructions where applicable. 
LLVM is then used to JIT compile the generated IR to executable code. The following sections describe each level of the IR and their lowering in more detail.

@@

%\TODO{Should we describe the dialect's type system?}
\TODO{Kr : consider focusing on the example instead of verbose description of optimizations. This section can be short, descriptions can come in later sections.}
\subsection{High Level IR}
As a first step Treebeard parses the input and generates a single MLIR operation, \texttt{predictForest} that represents inference using the input model on a set of rows. 
At this level the operator simply contains a collection of binary trees. Two optimizations, tiling and tree ordering are applied at this level. The objective of tiling is to group nodes together so that the tree can be walked one tile at a time instead of one node at a time. We demonstrate that tiling can specialize the traversal of individual trees to either balance heavily skewed trees or proritize walks leading to higher probability leafs. Figure~\ref{ttile} shows examples of the tiling transformation. \TODO{kr: draw and explain example}. The objective of tree padding and ordering is to group identically structured trees so that they can share the same traversal code. The predictForest function is now lowered to MIR with separate loop nests generated for traversing a group of identical trees. The generated MIR code is as shown below. 
\TODO{show unoptimized loop nest as a listing}.

%The operation contains within it a tree based representation of the model that can be manipulated by optimizing transformations. Transformations on the model such as tiling, tree reordering and leaf padding are done at this level. The structure of the loop nest to walk the iteration space of trees and inputs is also decided at this level of the IR. \TODO{Should we mention that there is a scheduling language to decide this?}

@@

\subsection{Mid Level IR}
The Mid Level IR optimizes the loop structures and tree walks. Firstly, the order in which the iteration space of trees and inputs is walked is determined and by re-ordering the loop nesting if necessary. Also, operations such as \texttt{isLeaf}, \texttt{traverseTile}, \texttt{getLeafValue} are introduced so that the traversal of trees explicitly represented. The following listing shows the IR for inference using a model with four trees on an input batch with two rows. The listed IR walks all trees for one input row before moving to the next row. One important point to note here is that details such as the data structure used for the trees are not explicitly encoded in the IR. This allows us to reuse optimization and lowering passes on this level of the IR regardless of what the final in memory representation of the model is.
